---
title: "Maternal Microbiome"
output: html_notebook
---

Introduction:

Our dataset contains levels of gut bacteria in healthy and unhealthy pregnant women. In this project, we are trying to find "candidate" bacteria -> bacteria that likely contributes to the health of the sample, so we can zero in on what might be causing illness.

Reading & Cleaning Data:

- This data contains 7 healthy and 7 unhealthy samples from maternal microbiomes

- To clean the data, we will remove the column containing averages for each bacteria and merge the first three columns into row names that contain the names for each bacteria
```{r}
my_data<- read.csv("./maternalmicrobiome.csv", header=TRUE)
my_data<- my_data[,1:17]
head(my_data)
```
```{r}
sub_data<- my_data[, 1:3]
row_names<-apply(sub_data, 1, function(r){paste(r[1], r[2], r[3], sep="_")})
row.names(my_data)<- row_names
my_data<- my_data[,-c(1, 2, 3)]
head(my_data)
```


Reformatting Data
```{r}
transposed_data<- data.frame(t(my_data))
head(transposed_data)
```
Finding candidates out of bacteria:

- We will use two sample t-tests to identify which bacteria are potential candidates, by looking at the bacteria with the most obvious difference in healthy and unhealthy samples

- The p-value is the probability that the differences in bacteria levels in these two groups are due to sampling errors

- The lower the p-value, the more likely the bacteria is a candidate
```{r}
candidates<- c()
significance_level<- 0.05
valid_p_values<- c()
for(i in 1:ncol(transposed_data)){
  healthy<-sapply(transposed_data[1:7, i], as.numeric)
  unhealthy<-sapply(transposed_data[8:14, i], as.numeric)
  results<- stats::t.test(healthy, unhealthy)
  if(results$p.value<=significance_level){
    candidates<- c(candidates, colnames(transposed_data)[i])
    valid_p_values<- c(valid_p_values, results$p.value)
  }
}
candidates
valid_p_values
```
With Bonferoni Correction:
```{r}
candidates_bon<- c()
significance_level_bon<- 0.05
valid_p_values_bon<- c()
for(i in 1:ncol(transposed_data)){
  healthy<-sapply(transposed_data[1:7, i], as.numeric)
  unhealthy<-sapply(transposed_data[8:14, i], as.numeric)
  results<- stats::t.test(healthy, unhealthy)
  significance_level_bon<-significance_level_bon/i
  if(results$p.value<=significance_level_bon){
    candidates_bon<- c(candidates_bon, colnames(transposed_data)[i])
    valid_p_values_bon<- c(valid_p_values_bon, results$p.value)
  }
}
candidates_bon
valid_p_values_bon
```

FDR Correction:
```{r}
significance_level<- 0.2
fdr_bacteria<-c()
fdr_p_vals<- c()
for(i in 1:ncol(transposed_data)){
  healthy<-sapply(transposed_data[1:7, i], as.numeric)
  unhealthy<-sapply(transposed_data[8:14, i], as.numeric)
  results<- stats::t.test(healthy, unhealthy)
  fdr_bacteria<- c(fdr_bacteria, colnames(transposed_data)[i])
  fdr_p_vals<- c(fdr_p_vals, results$p.value)
}
m<- data.frame(fdr_bacteria, fdr_p_vals, p.adjust(fdr_p_vals, "fdr"), p.adjust(fdr_p_vals, "bonferroni"))
names(m) <- c("bacteria_name", "p_values", "fdr_corrected_p_values", "bonferroni_corrected_p_values")

m[m[fdr_corrected_p_values]<0.05, ]
```
Summary:

- There are 3 bacteria with unadjusted p values less than are significance level of 0.05

- After adjusting the p values using bonferroni correction and fdr correction, there are no bacteria that still have adjusted p values under the significance level

- Our correction methods are too aggressive for our data

Assembling Data:

- The dataframe "m" contains the unadjusted p value, fdr corrected p value, and bonferroni corrected p value for each bacteria
```{r}
attach(m)
m<- m[order(p_values),]
m[1:30,]
```
Graphing Data

```{r}
plot_bacteria <- function(bacteria){
  boxplot(transposed_data[1:7, bacteria], transposed_data[8:14, bacteria], main=paste(bacteria), col=c("green", "red"), names=c("healthy", "unhealthy"), cex.main=0.7)
}
par(mfrow=c(2, 2))
plot_bacteria(m[1,1])
plot_bacteria(m[2,1])
plot_bacteria(m[3,1])
plot_bacteria(m[4,1])

```
```{r}
pdf("Test.pdf")
par(mfrow=c(4, 2))
top_7 <- m[1:7, 1]
for (i in 1:7){
  plot_bacteria(top_7[i])
}
graphics.off()
```
```{r}
barplot(sapply(m[, 2:4], mean), main="Average P Values", names.arg = c("Unadjusted", "FDR Corrected", "Bonferroni Corrected"), col=c("white", "grey", "black"))
```
Making an XGBoost Model:

- None of the bacteria had significant p-values, so our next step is using XGBoost to identify candidate bacteria

- xgboost stands for extreme gradient boosting, which is an ensemble learner that uses gradient boosting

- the gradient minimizes the loss from each model, and boosting is used to allow models to learn from the mistakes of previous models

- We'll use xgboost to make a model that receives a sample and predicts whether it is healthy or unhealthy based on its bacteria levels

- The important features of the model will reveal which bacteria are the strongest indicators of whether or not the sample is healthy

- Max.depth = 4

- eta = 1

- nrounds = 3

- objective = "binary:logistic"

https://www.r-bloggers.com/2018/05/tuning-xgboost-in-r-part-i/
https://www.analyticsvidhya.com/blog/2016/03/complete-guide-parameter-tuning-xgboost-with-codes-python/

```{r}
labels<- c(1, 1, 1, 1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0)
my_data<- transposed_data
mod <- xgboost(data=as.matrix(my_data), label=labels, max.depth=4, eta=1, nrounds=3,  objective = "binary:logistic")

```
Now we can look at which features (bacteria) our model is labelling as important
```{r}
importance<- xgb.importance(model = mod)
print(importance)
xgb.plot.importance(importance_matrix = importance, col="black", cex=0.4)

```
- The model has found that these four bacteria are the best indicators of health, even though these bacteria don't have the lowest p values
```{r}
pdf("SignificantBacteria.pdf")
par(mfrow=c(5, 2))
top_7 <- m[1:7, 1]
for (i in 1:7){
  plot_bacteria(top_7[i])
}
plot_bacteria(importance[1, Feature])
#plot_bacteria(importance[2, Feature])
plot_bacteria(importance[3, Feature])
plot_bacteria(importance[4, Feature])
graphics.off()

```

We can use clustering to determine whether or not the important features found by our model are able to separate healthy from unhealthy samples

We will use K Means clustering and hierarichical clustering.

When using K means clustering, we are able to specify that we want the variables to separate our data into 2 clusters, which in our case is healthy and unhealthy
```{r}
xgb_bacteria <- as.matrix(transposed_data[, importance[,Feature]])
#xgb_bacteria<- t(xgb_bacteria)
```
```{r}
fit<- kmeans(xgb_bacteria, 2)
fit$cluster
```
From the labels above, we can see that the kmeans clustering algorithm formed one cluster made up of 2 unhealthy samples,  and another cluster made up of the rest.

This indicates that the features our model labelled at important do not perform strongly in distinguishing healthy from unhealthy samples.

When using hierarchical clustering, the function determines how many clusters it should form
```{r}
d <- dist(xgb_bacteria, method = "euclidean") # distance matrix
fit <- hclust(d, method="ward.D")
plot(fit) # display dendogram
groups <- cutree(fit, k=2)
rect.hclust(fit, k=2, border="red")
```
- The graph above shows that even without our input, the function identified two clusters with the same split of samples.

- We are also able to see the minor distinctions the algorithm is making between healthy and unhealthy

- You can see that at a certain branch, a group of 4 unhealthy samples are separated by a group of 4 healthy samples

Clustering with low p value bacteria:

- I am interested to see if the 3 bacteria with the lowest p values from above will do a better job at clustering
```{r}
#head(transposed_data)
t_bacteria <- as.matrix(transposed_data[,m[1:3, 1]])
d <- dist(t_bacteria, method = "euclidean") # distance matrix
fit2 <- hclust(d, method="ward.D")
plot(fit2) # display dendogram
groups <- cutree(fit2, k=2)
rect.hclust(fit2, k=2, border="red")
```
- The bacteria with low p values were able to separate our data into a group of 4 unhealthy samples and another group containing the rest of the samples



Conclusion:

Due to our relatively small sample size, we were unable to find bacteria that are powerful indicators of whether or not a patient is healthy. However, we have identified bacteria of interest through t tests and xgboost, which could be interesting to research. The main bacteria of interest are:

- "Streptococcaceae.g_Streptococcus.s_anginosus" (low p value)

- "Lachnospiraceae.g_Blautia.s_wexlerae" (low p value)

- "Lachnospiraceae.g_Dorea.s_formicigenerans" (low p value and important feature)

- "Bacteroidaceae.g_Bacteroides.s_massiliensis"	(important feature)

From clustering, we found that the variables with low p values are better at distinguishing healthy from unhealthy samples, so the first 3 bacteria above are likely more significant



```{r}
test_data<- read.csv("./zymo_test_set.csv")
test_data<- test_data[-1,-c(1, 2, 3, 4, 5)]
dim(test_data)
```
```{r}
row_names<-apply(test_data, 1, function(r){paste(r[1], r[2], r[3], sep="_")})
rownames(test_data) <- row_names
test_data<- test_data[,-c(1, 2, 3)]
head(test_data)
```

```{r}
library(stringr)
s <- c("hello;world", "by;.world")
s<- sapply(s, function(e){as.character(str_replace(e, ";", "."))})
#str_replace("hello;world",";", ".")
s <- as.character(s)
s
```
```{r}

row_names<- as.character(row_names)

#transposed_data[, "Methanobacteriaceae;g_Methanobrevibacter;s_smithii"]
#test_data[      "Lachnospiraceae.g_Dorea.s_formicigenerans",]
which(row_names=="Lachnospiraceae;g_Dorea;s_formicigenerans")

test_data[201, ]
```
```{r}
row_names <- as.character(sapply(row_names, function(e){str_replace(e, ";", ".")}))
rownames(test_data)<- row_names
head(test_data)
```
```{r}
i <- intersect(row_names, colnames(transposed_data))
length(i)
```
```{r}
l_names<- c()
for(i in 1:length(row_names)){
  if (startsWith(row_names[i], "L")){
    l_names<- c(l_names, row_names[i])
  }
}
colnames(transposed_data)
```
```{r}
transposed_test<- data.frame(t(test_data))
full_data<- bind_rows(transposed_data, transposed_test)
full_data
```
```{r}
pos<- c(which(colnames(full_data)=="Lachnospiraceae.g_Dorea.s_formicigenerans"), which(colnames(full_data)=="Streptococcaceae.g_Streptococcus.s_anginosus"), which(colnames(full_data)=="Lachnospiraceae.g_Blautia.s_wexlerae"), which(colnames(full_data)=="Bacteroidaceae.g_Bacteroides.s_massiliensis"))
significant_rows <- full_data[, pos]
significant_rows
```
```{r}
healthy_train<- significant_rows[1:7,]
unhealthy_train<- significant_rows[8:14,]
healthy_test<- significant_rows[15:23,]
unhealthy_test<- significant_rows[23,]
```


```{r}
plot_train_and_test <- function(i){
  boxplot(healthy_train[,i], unhealthy_train[,i], healthy_test[,i], unhealthy_test[,i], names=c("Healthy\nTrain", "Unhealthy\nTrain", "Healthy\nTest", "Unhealthy\nTest"), main=colnames(significant_rows)[i])
}
#par(mfrow=c(2, 2))
plot_train_and_test(1)
plot_train_and_test(2)
plot_train_and_test(3)
plot_train_and_test(4)

```

```


